

## Section 2. Data Wrangling

I examine the data in an attempt to spot anomalies that need to be cleaned, as well as patterns and trends that exist in the data.


### 2.1. Data Summary and Structure

The table below provides summary statistics about the variables in the data frame.

```{r}
cpi_use %>% 
  skim() %>% 
  kbl() %>% 
  kable_material(c("striped", "hovered"),
                 html_font = "calibri") %>% 
  scroll_box(width = "100%", height = "300px")
```  


### 2.2. Missing Data 

The summary table shows that the data has some missing are missing values. I use the plot below to visualize where the missing values are located to try to understand how to deal with them.

```{r}
cpi_use %>% missmap(main = "Observed vs Missing Data")
```

The chart above shows that, overall, about 15 percent of the rows are have missing values. However, most of the missing data are from earlier years when BLS did not collect CPI data for those variables. Those missing values do not need to be replaced or removed as they will not affect our analysis in any way.

From the chart, it appears there are no recent data for `Motor Vehicle Insurance`. It appears that the BLS has discontinued the CPI series for the item as of the end of 2021. Hence, I remove the CPI series for `Motor Vehicle Insurance` from the data in the section below.


### 2.3. Creating and Dropping Variables

In this section, I remove `Motor Vehicle Insurance` from the data as explained above. I also drop the `period` column as it is not needed for this analysis.

Finally, I change the data type for the `year` field from character to number and create a new field called `date`, which combines the `month` and `year` columns.

```{r, message=FALSE}
cpi_use <- 
  cpi_use %>%
  select(-period, -motor.vehicle.insurance) %>% 
  mutate(year = as.numeric(year),
         date = as.POSIXct(strptime(paste0('1',cpi_use$periodName,cpi_use$year),'%d%B%Y'))
         )
```


## Section 3. Inflation in NY-Metro Area

In calculating inflation, I do a year-over-year comparison since the CPI data used in this analysis are not seasonally adjusted. I compare CPI for the month to CPI for the same month of the previous year.


### 3.1. Monthly Inflation

The code chunk below calculates 12-month percent change in CPI per month for all the components and stores that in a new data frame called `monthly_pchg`.

```{r}
monthly_pchg <- cpi_use %>% 
  arrange(date) %>%
  mutate(year = as.character(year)) %>% 
  mutate_if(is.numeric, 
            ~round((./lag(., 12)-1), 3)) %>% 
  filter(year != "2003")
```

The table below shows the monthly inflation for all components.

```{r, echo=FALSE}
monthly_pchg %>% 
  mutate(year = as.numeric(year)) %>% 
  filter(year == max(year)) %>%
  kbl() %>% 
  kable_classic_2(c("striped", "hovered"),
                      html_font = "calibri") %>% 
  scroll_box(width = "100%", height = "300px")
```


### 3.2. Inflation for the Latest Month {.tabset .tabset-pills .tabset-fade}

In this section, I use bar charts to depict inflation in the NY-Metro area for the current month - December 2022.

The chart in the first tab depicts headline and core inflation, while the second tab shows inflation for some of the important CPI components.


#### Headline & Core Inflation

```{r}
monthly_pchg %>% select(date, all.items, all.items.less.food.energy) %>% 
  filter(date == max(date)) %>% 
  rename('Headline' = all.items,
         'Core' = all.items.less.food.energy) %>% 
  pivot_longer(cols = 2:3,
               names_to =  "Item",
               values_to =  "p_chg",
               values_drop_na =  T) %>%
  ggplot(aes(x = Item, y = p_chg)) +
  geom_bar(position = 'dodge', fill = '#bb969a', stat = "identity") +
  scale_y_continuous(labels = percent) +
  geom_text(aes(label = percent(p_chg, accuracy = .1))) +
  theme_economist() +
  theme(panel.grid = element_blank()) +
  labs(x = NULL,
       y = "Inflation Rate (%)",
       title = "YoY Change in Inflation, August 2023",
       subtitle = "Healine and Core CPI",
       caption = "Non seasonally adjusted data")
```

#### Select Items

```{r}
monthly_pchg %>% 
  select(date, food, rent.of.primary.residence,household.energy, 
         medical.care, apparel, tuition.other.school.fees.childcare,
         recreation, new.used.motor.vehicles, gasoline.all.types, energy) %>%
  rename('Food' = food,
         'Rent' = rent.of.primary.residence,
         'Home utilities' = household.energy,
         'Medical Care' = medical.care,
         'Apparel' = apparel,
         'Tuition & Child Care' = tuition.other.school.fees.childcare,
         'Recreation' = recreation,
         'Vehicles' = new.used.motor.vehicles,
         'Gasoline' = gasoline.all.types,
         'Energy' = energy) %>% 
  filter(date == max(date)) %>% 
  pivot_longer(cols = 2:11,
               names_to = "item",
               values_to = "p_chg",
               values_drop_na = T) %>% 
  ggplot(aes(x = reorder(item, -p_chg),
             y = p_chg)
  ) +
  geom_bar(position='dodge', fill = '#bb969a', stat='identity') +
  #facet_wrap(~item) +
  scale_y_continuous(labels = percent) +
  geom_text(aes(label = percent(p_chg, accuracy = .1))) +
  coord_flip() +
  theme_economist() +
  theme(panel.grid.major = element_blank()) +
  labs(x = NULL,
       y = "Inflation Rate (%)",
       title = "YoY Change in Prices, Dec. 2022",
       subtitle = "Select CPI Prices",
       caption = "Not seasonally adjusted")
```


### 3.3. Trends in Inflation {.tabset .tabset-pills .tabset-fade}

In this section, I use line graphs to depict the trend in monthly inflation, highlighting the recent spikes in the prices of certain items.

#### Headline & Core

```{r}
monthly_pchg %>% select(date, all.items, all.items.less.food.energy) %>% 
  rename('Headline' = all.items,
         'Core' = all.items.less.food.energy) %>%
  pivot_longer(cols = 2:3, names_to = "Component:",
               values_to = "p_chg", values_drop_na = T) %>%
  ggplot(aes(x = date, y = p_chg)) +
  geom_line(aes(color = `Component:`), lwd = 1) +
  scale_y_continuous(labels = percent) +
  scale_color_manual(values = c("#bb969a", "#00578a")) +
  geom_hline(yintercept = 0, color = "White") +
  theme_economist() +
  theme(panel.grid.major = element_blank()) +
  labs(x = NULL,
       y = "Inflation Rate (%)",
       title = "YoY Change in Prices",
       subtitle = "Headline & Core Inflation",
       caption = "Not seasonally adjusted")

```


#### Gasoline

```{r}
monthly_pchg %>% select(date, gasoline.unleaded.regular, gasoline.unleaded.midgrade,
                        gasoline.unleaded.premium) %>% 
  rename(Regular = gasoline.unleaded.regular,
         Midgrade = gasoline.unleaded.midgrade,
         Premium = gasoline.unleaded.premium) %>%
  pivot_longer(cols = 2:4, names_to = "Component:",
               values_to = "p_chg", values_drop_na = T) %>%
  ggplot(aes(x = date, y = p_chg)) +
  geom_line(aes(color = `Component:`), lwd = 1) +
  geom_hline(yintercept = 0, color = "White") +
  scale_y_continuous(labels = percent) +
  scale_color_manual(values = c("#ff8181" ,"#bb969a", "#00578a")) +
  theme_economist() +
  theme(panel.grid.major = element_blank()) +
  labs(x = NULL,
       y = "Inflation Rate (%)",
       title = "YoY Change in Gas Prices",
       caption = "Not seasonally adjusted")

```


#### Food

```{r}
monthly_pchg %>%
  select(date, food.at.home, food.away.from.home) %>% 
  rename(`Food at Home` = food.at.home, `Food From Outside` = food.away.from.home) %>% 
    pivot_longer(cols = 2:3, names_to = "Component:",
               values_to = "p_chg", values_drop_na = T) %>%
  ggplot(aes(x = date, y = p_chg)) +
  geom_line(aes(color = `Component:`), lwd = 1) +
  geom_hline(yintercept = 0, color = "White") +
  scale_y_continuous(labels = percent) +
  scale_color_manual(values = c("#bb969a", "#00578a")) +
  theme_economist() +
  theme(panel.grid.major = element_blank()) +
  labs(x = NULL,
       y = "Inflation Rate (%)",
       title = "YoY Change in Food Prices",
       subtitle = "Food at Home vs. Food From Outside",
       caption = "Not seasonally adjusted")
```


#### Select Food Items

```{r}
monthly_pchg %>%
  select(date, meats.poultry.fish.egg, fruits.vegetables, 
         cereals.bakery.products, dairy.related.products) %>% 
  rename(`Meat,Poultry,Fish,Eggs` = meats.poultry.fish.egg, `Fruits & Veggies` = fruits.vegetables, 
         `Cereal & Baked Goods` = cereals.bakery.products, Dairy = dairy.related.products) %>% 
    pivot_longer(cols = 2:5, names_to = "Component:",
               values_to = "p_chg", values_drop_na = T) %>%
  ggplot(aes(x = date, y = p_chg)) +
  geom_line(aes(color = `Component:`), lwd = 1) +
  geom_hline(yintercept = 0, color = "White") +
  scale_y_continuous(labels = percent) +
  scale_color_manual(values = c("#ff8181" ,"#bb969a", "#00578a", "#ff8c00")) +
  theme_economist() +
  theme(panel.grid.major = element_blank()) +
  labs(x = NULL,
       y = "Inflation Rate (%)",
       title = "YoY Change in Food Prices",
       subtitle = "Select Food Items",
       caption = "Not seasonally adjusted")
```


#### Shelter

```{r}
monthly_pchg %>%
  select(date, rent.of.primary.residence, owners.equivalent.rent.of.primary.residence) %>% 
  rename(Rent = rent.of.primary.residence,
         Mortgage = owners.equivalent.rent.of.primary.residence) %>% 
    pivot_longer(cols = 2:3, names_to = "Component:",
               values_to = "p_chg", values_drop_na = T) %>%
  ggplot(aes(x = date, y = p_chg)) +
  geom_line(aes(color = `Component:`), lwd = 1) +
  geom_hline(yintercept = 0, color = "White") +
  scale_y_continuous(labels = percent) +
  scale_color_manual(values = c("#bb969a", "#00578a")) +
  theme_economist() +
  theme(panel.grid.major = element_blank()) +
  labs(x = NULL,
       y = "Inflation Rate (%)",
       title = "YoY Change in Shelter Prices",
       caption = "Not seasonally adjusted")
```


#### Vehicles

```{r}
monthly_pchg %>%
  select(date, new.vehicles, used.cars.trucks) %>% 
  rename(`New Vehicle` = new.vehicles,
         `Used Vehicle` = used.cars.trucks) %>% 
    pivot_longer(cols = 2:3, names_to = "Component:",
               values_to = "p_chg", values_drop_na = T) %>%
  ggplot(aes(x = date, y = p_chg)) +
  geom_line(aes(color = `Component:`), lwd = 1) +
  geom_hline(yintercept = 0, color = "White") +
  scale_y_continuous(labels = percent) +
  scale_color_manual(values = c("#bb969a", "#00578a")) +
  theme_economist() +
  theme(panel.grid.major = element_blank()) +
  labs(x = NULL,
       y = "Inflation Rate (%)",
       title = "YoY Change in Vehicle Prices",
       caption = "Not seasonally adjusted")
```


