



## Section 2. Exploratory Data Analysis

Here, I examine the data to spot anomalies that need to be cleaned, as well as patterns and trends that exist in the data.

### 2.1. Data Summary and Structure

The table below shows a summary of the data and distributions of  some of its variables.

```{r}
cpi_use %>% 
  skim() %>% 
  kbl() %>% 
  kable_material(c("striped", "hovered"),
                 html_font = "calibri") %>% 
  scroll_box(width = "100%", height = "300px")
```  


### 2.2. Missing Data 

The summary table shows that there are missing values in the data. I use the plot below to visualize where the missing values are located to try to understand how to deal with them.

```{r}
cpi_use %>% missmap(main = "Observed vs Missing Data")
```

The chart above shows that, overall, about 15 percent of rows are have missing values. However,most of the missing data are from earlier years when BLS did not collect CPI data for those variables. Those missing values do not need to be replaced or removed as they will not affect our analysis in any way.

The chart and table above, it appears there are no recent data for `Motor Vehicle Insurance`. It appears that the BLS has discontinued the CPI series for the component as of the end of 2021. Hence, I delete all data for that CPI component during the cleaning process.


### 2.3. Data Wrangling

Here, I remove the `Motor Vehicle Insurance` column convert the `year` column from to numeric. I also create a `date` column, which combines the month and year colums.

```{r, message=FALSE}
cpi_use <- 
  cpi_use %>%
  select(-period, -motor.vehicle.insurance) %>% 
  mutate(year = as.numeric(year),
         date = as.POSIXct(strptime(paste0('1',cpi_use$periodName,cpi_use$year),'%d%B%Y'))
         )
```


## Section 3. Inflation in NY-Metro Area

Ater exploring and cleaning the data, I examine the monthly inflation in the NY-Metro Area. Since I AM using a non-seasonally adjusted data, I do a year-over-year percentage change, comparing CPI for a month to CPI for the same month of the previous year.


### 3.1. Monthly Inflation for All Components 

Let's calculate the 12-month percent change in CPI per month for all the components and store that in a new data frame (`monthly_pchg`).

```{r}
monthly_pchg <- cpi_use %>% 
  arrange(date) %>%
  mutate(year = as.character(year)) %>% 
  mutate_if(is.numeric, 
            ~round((./lag(., 12)-1), 3)) %>% 
  filter(year != "2003")
```

Here is a table that shows the y-o-y percent changes in prices for all months.

```{r, echo=FALSE}
monthly_pchg %>% 
  mutate(year = as.numeric(year)) %>% 
  filter(year == max(year)) %>%
  kbl() %>% 
  kable_classic_2(c("striped", "hovered"),
                      html_font = "calibri") %>% 
  scroll_box(width = "100%", height = "300px")
```


### 3.2. Inflation for the Latest Month {.tabset .tabset-pills .tabset-fade}

In this section, I analyze inflation for the most recent month for which data is available - December 2022.

The chart in the first depicts headline and core inflation for December 2022 in the NY-Metro Area. 

The second tab shows inflation for the month for some of important CPI components that matter in the NY-Metro Area and the rest of the U.S.


#### Headline & Core Inflation

```{r}
monthly_pchg %>% select(date, all.items, all.items.less.food.energy) %>% 
  filter(date == max(date)) %>% 
  rename('Headline' = all.items,
         'Core' = all.items.less.food.energy) %>% 
  pivot_longer(cols = 2:3,
               names_to =  "Item",
               values_to =  "p_chg",
               values_drop_na =  T) %>%
  ggplot(aes(x = Item, y = p_chg)) +
  geom_bar(position = 'dodge', fill = '#bb969a', stat = "identity") +
  scale_y_continuous(labels = percent) +
  geom_text(aes(label = percent(p_chg, accuracy = .1))) +
  theme_economist() +
  theme(panel.grid = element_blank()) +
  labs(x = NULL,
       y = "Inflation Rate (%)",
       title = "YoY Change in Inflation, Dec. 2022",
       subtitle = "Healine and Core CPI",
       caption = "Non seasonally adjusted data")
```

#### Select Items

```{r}
monthly_pchg %>% 
  select(date, food, rent.of.primary.residence,household.energy, 
         medical.care, apparel, tuition.other.school.fees.childcare,
         recreation, new.used.motor.vehicles, gasoline.all.types, energy) %>%
  rename('Food' = food,
         'Rent' = rent.of.primary.residence,
         'Home utilities' = household.energy,
         'Medical Care' = medical.care,
         'Apparel' = apparel,
         'Tuition & Child Care' = tuition.other.school.fees.childcare,
         'Recreation' = recreation,
         'Vehicles' = new.used.motor.vehicles,
         'Gasoline' = gasoline.all.types,
         'Energy' = energy) %>% 
  filter(date == max(date)) %>% 
  pivot_longer(cols = 2:11,
               names_to = "item",
               values_to = "p_chg",
               values_drop_na = T) %>% 
  ggplot(aes(x = reorder(item, -p_chg),
             y = p_chg)
  ) +
  geom_bar(position='dodge', fill = '#bb969a', stat='identity') +
  #facet_wrap(~item) +
  scale_y_continuous(labels = percent) +
  geom_text(aes(label = percent(p_chg, accuracy = .1))) +
  coord_flip() +
  theme_economist() +
  theme(panel.grid.major = element_blank()) +
  labs(x = NULL,
       y = "Inflation Rate (%)",
       title = "YoY Change in Prices, Dec. 2022",
       subtitle = "Select CPI Prices",
       caption = "Not seasonally adjusted")
```


### 3.3. Trends in Inflation {.tabset .tabset-pills .tabset-fade}

In this section, I use line graphs to depict the trend in monthly inflation to show the recent spikes in the prices of certain items.

#### Headline & Core

```{r}
monthly_pchg %>% select(date, all.items, all.items.less.food.energy) %>% 
  rename('Headline' = all.items,
         'Core' = all.items.less.food.energy) %>%
  pivot_longer(cols = 2:3, names_to = "Component:",
               values_to = "p_chg", values_drop_na = T) %>%
  ggplot(aes(x = date, y = p_chg)) +
  geom_line(aes(color = `Component:`), lwd = 1) +
  scale_y_continuous(labels = percent) +
  scale_color_manual(values = c("#bb969a", "#00578a")) +
  geom_hline(yintercept = 0, color = "White") +
  theme_economist() +
  theme(panel.grid.major = element_blank()) +
  labs(x = NULL,
       y = "Inflation Rate (%)",
       title = "YoY Change in Prices",
       subtitle = "Headline & Core Inflation",
       caption = "Not seasonally adjusted")

```


#### Gasoline

```{r}
monthly_pchg %>% select(date, gasoline.unleaded.regular, gasoline.unleaded.midgrade,
                        gasoline.unleaded.premium) %>% 
  rename(Regular = gasoline.unleaded.regular,
         Midgrade = gasoline.unleaded.midgrade,
         Premium = gasoline.unleaded.premium) %>%
  pivot_longer(cols = 2:4, names_to = "Component:",
               values_to = "p_chg", values_drop_na = T) %>%
  ggplot(aes(x = date, y = p_chg)) +
  geom_line(aes(color = `Component:`), lwd = 1) +
  geom_hline(yintercept = 0, color = "White") +
  scale_y_continuous(labels = percent) +
  scale_color_manual(values = c("#ff8181" ,"#bb969a", "#00578a")) +
  theme_economist() +
  theme(panel.grid.major = element_blank()) +
  labs(x = NULL,
       y = "Inflation Rate (%)",
       title = "YoY Change in Gas Prices",
       caption = "Not seasonally adjusted")

```


#### Food

```{r}
monthly_pchg %>%
  select(date, food.at.home, food.away.from.home) %>% 
  rename(`Food at Home` = food.at.home, `Food From Outside` = food.away.from.home) %>% 
    pivot_longer(cols = 2:3, names_to = "Component:",
               values_to = "p_chg", values_drop_na = T) %>%
  ggplot(aes(x = date, y = p_chg)) +
  geom_line(aes(color = `Component:`), lwd = 1) +
  geom_hline(yintercept = 0, color = "White") +
  scale_y_continuous(labels = percent) +
  scale_color_manual(values = c("#bb969a", "#00578a")) +
  theme_economist() +
  theme(panel.grid.major = element_blank()) +
  labs(x = NULL,
       y = "Inflation Rate (%)",
       title = "YoY Change in Food Prices",
       subtitle = "Food at Home vs. Food From Outside",
       caption = "Not seasonally adjusted")
```


#### Select Food Items

```{r}
monthly_pchg %>%
  select(date, meats.poultry.fish.egg, fruits.vegetables, 
         cereals.bakery.products, dairy.related.products) %>% 
  rename(`Meat,Poultry,Fish,Eggs` = meats.poultry.fish.egg, `Fruits & Veggies` = fruits.vegetables, 
         `Cereal & Baked Goods` = cereals.bakery.products, Dairy = dairy.related.products) %>% 
    pivot_longer(cols = 2:5, names_to = "Component:",
               values_to = "p_chg", values_drop_na = T) %>%
  ggplot(aes(x = date, y = p_chg)) +
  geom_line(aes(color = `Component:`), lwd = 1) +
  geom_hline(yintercept = 0, color = "White") +
  scale_y_continuous(labels = percent) +
  scale_color_manual(values = c("#ff8181" ,"#bb969a", "#00578a", "#ff8c00")) +
  theme_economist() +
  theme(panel.grid.major = element_blank()) +
  labs(x = NULL,
       y = "Inflation Rate (%)",
       title = "YoY Change in Food Prices",
       subtitle = "Select Food Items",
       caption = "Not seasonally adjusted")
```


#### Shelter

```{r}
monthly_pchg %>%
  select(date, rent.of.primary.residence, owners.equivalent.rent.of.primary.residence) %>% 
  rename(Rent = rent.of.primary.residence,
         Mortgage = owners.equivalent.rent.of.primary.residence) %>% 
    pivot_longer(cols = 2:3, names_to = "Component:",
               values_to = "p_chg", values_drop_na = T) %>%
  ggplot(aes(x = date, y = p_chg)) +
  geom_line(aes(color = `Component:`), lwd = 1) +
  geom_hline(yintercept = 0, color = "White") +
  scale_y_continuous(labels = percent) +
  scale_color_manual(values = c("#bb969a", "#00578a")) +
  theme_economist() +
  theme(panel.grid.major = element_blank()) +
  labs(x = NULL,
       y = "Inflation Rate (%)",
       title = "YoY Change in Shelter Prices",
       caption = "Not seasonally adjusted")
```


#### Vehicles

```{r}
monthly_pchg %>%
  select(date, new.vehicles, used.cars.trucks) %>% 
  rename(`New Vehicle` = new.vehicles,
         `Used Vehicle` = used.cars.trucks) %>% 
    pivot_longer(cols = 2:3, names_to = "Component:",
               values_to = "p_chg", values_drop_na = T) %>%
  ggplot(aes(x = date, y = p_chg)) +
  geom_line(aes(color = `Component:`), lwd = 1) +
  geom_hline(yintercept = 0, color = "White") +
  scale_y_continuous(labels = percent) +
  scale_color_manual(values = c("#bb969a", "#00578a")) +
  theme_economist() +
  theme(panel.grid.major = element_blank()) +
  labs(x = NULL,
       y = "Inflation Rate (%)",
       title = "YoY Change in Vehicle Prices",
       caption = "Not seasonally adjusted")
```


