
## Introduction

Inflation has been a major topic of discussion over the past few years. The COVID-19 pandemic and the shutdowns that ensued from it brought on global supply shortages, causing prices of goods and services to skyrocket across industries. Notable among them was the surge in oil prices, which was initially caused low investment in oil production prior to the pandemic and exacerbated by the Russia-Ukraine conflict.

The goal of this project is to track inflation in the New York Metro Area using monthly consumer price index (CPI) data from the U.S. Bureau of Labor Statistics (BLS). It calculates the 12-month percent change in prices for CPI components and uses visuals to highlights headline and core inflation, as well as price changes for select components. Lastly, using an ARIMA model, it predicts the monthly headline inflation in the NY-Metro Area for the next 12 months.

## Section 1. Download CPI Data using BLS API {.tabset .tabset-fade .tabset-pills}

This analysis uses the non-seasonally adjusted CPI data for the NY-Metro Area published by the BLS. The data is extracted using an API, and then transformed, loaded and cleaned for use in this analysis.^[See `child 1.Rmd` file for the API used to extract and transform the data] The latest month for which CPI data is available is **December 2022**.

I extracted monthly data for all 61 CPI components for which BLS publishes data pertaining to the NY Metro Area. The extracted data were cleaned and merged into a data frame object called `cpi_use`.^[Since the BLS API only allows a maximum of 50 variables or items per API request, the download process is broken into two parts - first 50 variables and last 11 variables] The data includes monthly CPI for 2003 through 2022 as  BLS limits data downloads via API to 20 years per request. In the future, the project may be expanded to include data prior to 2003.


```{r, include=FALSE, echo=FALSE}

# Downloading the first 50 components going back to 2003.

payload <- list(
  'seriesid'=c(
    'CUURS12ASA0','CUURS12ASA0E','CUURS12ASA0L1E','CUURS12ASA0L2','CUURS12ASA0L5',
    'CUURS12ASA0LE','CUURS12ASAA','CUURS12ASAC','CUURS12ASACL1','CUURS12ASACL11',
    'CUURS12ASAD','CUURS12ASAE','CUURS12ASAEC','CUURS12ASAES','CUURS12ASAF',
    'CUURS12ASAF1','CUURS12ASAF11','CUURS12ASAF111','CUURS12ASAF112','CUURS12ASAF113',
    'CUURS12ASAF114','CUURS12ASAF115','CUURS12ASAF116','CUURS12ASAG','CUURS12ASAGC',
    'CUURS12ASAGS','CUURS12ASAH','CUURS12ASAH1','CUURS12ASAH2','CUURS12ASAH21',
    'CUURS12ASAH3','CUURS12ASAH31','CUURS12ASAM','CUURS12ASAN','CUURS12ASANL1',
    'CUURS12ASANL11','CUURS12ASAR','CUURS12ASAS','CUURS12ASASL2RS','CUURS12ASASL5',
    'CUURS12ASAT','CUURS12ASAT1','CUURS12ASATCLTB','CUURS12ASEEB','CUURS12ASEFJ',
    'CUURS12ASEFV','CUURS12ASEHA','CUURS12ASEHC','CUURS12ASEHC01','CUURS12ASEHF'
  ),
  'startyear'=2003,
  'endyear'=2023,
  'catalog'=FALSE,
  'calculations'=TRUE,
  'annualaverage'=FALSE,
  #'seriestitle'=TRUE,
  #'area'=TRUE,
  'registrationKey'='c798401369a64ada824b64aed51716ed')
response <- blsAPI(payload, 2)
json <- fromJSON(response)

## Process results
apiDF <- function(data){
  df <- data.frame(year=character(),
                   period=character(),
                   periodName=character(),
                   value=character(),
                   stringsAsFactors=FALSE)
  
  i <- 0
  for(d in data){
    i <- i + 1
    df[i,] <- unlist(d)
  }
  return(df)
}

all.items <- apiDF(json$Results$series[[1]]$data)
energy <- apiDF(json$Results$series[[2]]$data)
all.items.less.food.energy <- apiDF(json$Results$series[[3]]$data)
all.items.less.shelter <- apiDF(json$Results$series[[4]]$data)
all.items.less.medical.care <- apiDF(json$Results$series[[5]]$data)
all.items.less.energy <- apiDF(json$Results$series[[6]]$data)
apparel <- apiDF(json$Results$series[[7]]$data)
commodities <- apiDF(json$Results$series[[8]]$data)
commodities.less.food <- apiDF(json$Results$series[[9]]$data)
commodities.less.food.beverages <- apiDF(json$Results$series[[10]]$data)
durables <- apiDF(json$Results$series[[11]]$data)
education.communication <- apiDF(json$Results$series[[12]]$data)
education.communication.commodities <- apiDF(json$Results$series[[13]]$data)
education.communication.services <- apiDF(json$Results$series[[14]]$data)
food.beverages <- apiDF(json$Results$series[[15]]$data)
food <- apiDF(json$Results$series[[16]]$data)
food.at.home <- apiDF(json$Results$series[[17]]$data)
cereals.bakery.products <- apiDF(json$Results$series[[18]]$data)
meats.poultry.fish.egg <- apiDF(json$Results$series[[19]]$data)
fruits.vegetables <- apiDF(json$Results$series[[20]]$data)
nonalcoholic.beverages.beverage.materials <- apiDF(json$Results$series[[21]]$data)
other.food.at.home <- apiDF(json$Results$series[[22]]$data)
alcoholic.beverages <- apiDF(json$Results$series[[23]]$data)
other.goods.services <- apiDF(json$Results$series[[24]]$data)
other.goods <- apiDF(json$Results$series[[25]]$data)
other.personal.services <- apiDF(json$Results$series[[26]]$data)
housing <- apiDF(json$Results$series[[27]]$data)
shelter <- apiDF(json$Results$series[[28]]$data)
fuels.utilities <- apiDF(json$Results$series[[29]]$data)
household.energy <- apiDF(json$Results$series[[30]]$data)
household.furnishing.operations <- apiDF(json$Results$series[[31]]$data)
household.furnishing.supplies <- apiDF(json$Results$series[[32]]$data)
medical.care <- apiDF(json$Results$series[[33]]$data)
nondurables <- apiDF(json$Results$series[[34]]$data)
nondurables.less.food <- apiDF(json$Results$series[[35]]$data)
nondurables.less.food.beverages <- apiDF(json$Results$series[[36]]$data)
recreation <- apiDF(json$Results$series[[37]]$data)
services <- apiDF(json$Results$series[[38]]$data)
services.less.rent.of.shelter <- apiDF(json$Results$series[[39]]$data)
services.less.medical.services <- apiDF(json$Results$series[[40]]$data)
transportation <- apiDF(json$Results$series[[41]]$data)
private.transportation <- apiDF(json$Results$series[[42]]$data)
transportation.commodities.less.motor.fuel <- apiDF(json$Results$series[[43]]$data)
tuition.other.school.fees.childcare <- apiDF(json$Results$series[[44]]$data)
dairy.related.products <- apiDF(json$Results$series[[45]]$data)
food.away.from.home <- apiDF(json$Results$series[[46]]$data)
rent.of.primary.residence <- apiDF(json$Results$series[[47]]$data)
owners.equivalent.rent.of.residences <- apiDF(json$Results$series[[48]]$data)
owners.equivalent.rent.of.primary.residence <- apiDF(json$Results$series[[49]]$data)
energy.services <- apiDF(json$Results$series[[50]]$data)


## Rename value prior to merging
names(all.items)[4]<- 'all.items'
names(energy)[4]<- 'energy'
names(all.items.less.food.energy)[4]<- 'all.items.less.food.energy'
names(all.items.less.shelter)[4]<- 'all.items.less.shelter'
names(all.items.less.medical.care)[4]<- 'all.items.less.medical.care'
names(all.items.less.energy)[4]<- 'all.items.less.energy'
names(apparel)[4]<- 'apparel'
names(commodities)[4]<- 'commodities'
names(commodities.less.food)[4]<- 'commodities.less.food'
names(commodities.less.food.beverages)[4]<- 'commodities.less.food.beverages'
names(durables)[4]<- 'durables'
names(education.communication)[4]<- 'education.communication'
names(education.communication.commodities)[4]<- 'education.communication.commodities'
names(education.communication.services)[4]<- 'education.communication.services'
names(food.beverages)[4]<- 'food.beverages'
names(food)[4]<- 'food'
names(food.at.home)[4]<- 'food.at.home'
names(cereals.bakery.products)[4]<- 'cereals.bakery.products'
names(meats.poultry.fish.egg)[4]<- 'meats.poultry.fish.egg'
names(fruits.vegetables)[4]<- 'fruits.vegetables'
names(nonalcoholic.beverages.beverage.materials)[4]<- 'nonalcoholic.beverages.beverage.materials'
names(other.food.at.home)[4]<- 'other.food.at.home'
names(alcoholic.beverages)[4]<- 'alcoholic.beverages'
names(other.goods.services)[4]<- 'other.goods.services'
names(other.goods)[4]<- 'other.goods'
names(other.personal.services)[4]<- 'other.personal.services'
names(housing)[4]<- 'housing'
names(shelter)[4]<- 'shelter'
names(fuels.utilities)[4]<- 'fuels.utilities'
names(household.energy)[4]<- 'household.energy'
names(household.furnishing.operations)[4]<- 'household.furnishing.operations'
names(household.furnishing.supplies)[4]<- 'household.furnishing.supplies'
names(medical.care)[4]<- 'medical.care'
names(nondurables)[4]<- 'nondurables'
names(nondurables.less.food)[4]<- 'nondurables.less.food'
names(nondurables.less.food.beverages)[4]<- 'nondurables.less.food.beverages'
names(recreation)[4]<- 'recreation'
names(services)[4]<- 'services'
names(services.less.rent.of.shelter)[4]<- 'services.less.rent.of.shelter'
names(services.less.medical.services)[4]<- 'services.less.medical.services'
names(transportation)[4]<- 'transportation'
names(private.transportation)[4]<- 'private.transportation'
names(transportation.commodities.less.motor.fuel)[4]<- 'transportation.commodities.less.motor.fuel'
names(tuition.other.school.fees.childcare)[4]<- 'tuition.other.school.fees.childcare'
names(dairy.related.products)[4]<- 'dairy.related.products'
names(food.away.from.home)[4]<- 'food.away.from.home'
names(rent.of.primary.residence)[4]<- 'rent.of.primary.residence'
names(owners.equivalent.rent.of.residences)[4]<- 'owners.equivalent.rent.of.residences'
names(owners.equivalent.rent.of.primary.residence)[4]<- 'owners.equivalent.rent.of.primary.residence'

## Merge data frames, convert cpi columns to numeric and create column for date

data_list <- list(
  all.items,energy, all.items.less.food.energy, all.items.less.shelter, all.items.less.medical.care,
  all.items.less.energy, apparel, commodities, commodities.less.food, commodities.less.food.beverages,
  durables, education.communication, education.communication.commodities, education.communication.services,
  food.beverages, food, food.at.home, cereals.bakery.products, meats.poultry.fish.egg,
  fruits.vegetables, nonalcoholic.beverages.beverage.materials, other.food.at.home, alcoholic.beverages,
  other.goods.services, other.goods, other.personal.services, housing, shelter,
  fuels.utilities, household.energy, household.furnishing.operations, household.furnishing.supplies,
  medical.care, nondurables, nondurables.less.food, nondurables.less.food.beverages, recreation,
  services, services.less.rent.of.shelter, services.less.medical.services, transportation, private.transportation,
  transportation.commodities.less.motor.fuel, tuition.other.school.fees.childcare, dairy.related.products,
  food.away.from.home, rent.of.primary.residence, owners.equivalent.rent.of.residences, owners.equivalent.rent.of.primary.residence,
  energy.services
)

cpi <- data_list %>% reduce(merge, all=T) %>% 
  mutate_at(c(4:53), as.numeric)
## cpi <- Reduce(function(x,y) merge(x,y, all=T), data_list)         #alternative to line above for merging

## Create a new column for date
#cpi <- cpi %>% mutate(date = as.POSIXct(strptime(paste0('1',cpi$periodName,cpi$year),
#                                                 '%d%B%Y')))

#cpi$date <- as.Date(with(cpi, paste(year,periodName,'1',sep="-")),"%Y-%m-%d")

## Clean envirnonment
rm (
  all.items,energy, all.items.less.food.energy, all.items.less.shelter, all.items.less.medical.care,
  all.items.less.energy, apparel, commodities, commodities.less.food, commodities.less.food.beverages,
  durables, education.communication, education.communication.commodities, education.communication.services,
  food.beverages, food, food.at.home, cereals.bakery.products, meats.poultry.fish.egg,
  fruits.vegetables, nonalcoholic.beverages.beverage.materials, other.food.at.home, alcoholic.beverages,
  other.goods.services, other.goods, other.personal.services, housing, shelter,
  fuels.utilities, household.energy, household.furnishing.operations, household.furnishing.supplies,
  medical.care, nondurables, nondurables.less.food, nondurables.less.food.beverages, recreation,
  services, services.less.rent.of.shelter, services.less.medical.services, transportation, private.transportation,
  transportation.commodities.less.motor.fuel, tuition.other.school.fees.childcare, dairy.related.products,
  food.away.from.home, rent.of.primary.residence, owners.equivalent.rent.of.residences, owners.equivalent.rent.of.primary.residence,
  energy.services
)
```


```{r, include=FALSE, echo=FALSE}

# Downloading the last 11 components going back to 2003.

payload <- list('seriesid'=c(
    'CUURS12ASEHF01','CUURS12ASEHF02','CUURS12ASETA','CUURS12ASETA01','CUURS12ASETA02',
    'CUURS12ASETB','CUURS12ASETB01','CUURS12ASETE','CUURS12ASS47014','CUURS12ASS47015',
    'CUURS12ASS47016'
  ),
  'startyear'=2003,
  'endyear'=2023,
  'catalog'=FALSE,
  'calculations'=TRUE,
  'annualaverage'=FALSE,
  #'seriestitle'=TRUE,
  #'area'=TRUE,
  'registrationKey'='c798401369a64ada824b64aed51716ed')
response <- blsAPI(payload, 2)
json <- fromJSON(response)

## Process results
apiDF <- function(data){
  df <- data.frame(year=character(),
                   period=character(),
                   periodName=character(),
                   value=character(),
                   stringsAsFactors=FALSE)
  
  i <- 0
  for(d in data){
    i <- i + 1
    df[i,] <- unlist(d)
  }
  return(df)
}

electricity <- apiDF(json$Results$series[[1]]$data)
utility.piped.gas.services <- apiDF(json$Results$series[[2]]$data)
new.used.motor.vehicles <- apiDF(json$Results$series[[3]]$data)
new.vehicles <- apiDF(json$Results$series[[4]]$data)
used.cars.trucks <- apiDF(json$Results$series[[5]]$data)
motor.fuel <- apiDF(json$Results$series[[6]]$data)
gasoline.all.types <- apiDF(json$Results$series[[7]]$data)
motor.vehicle.insurance <- apiDF(json$Results$series[[8]]$data)
gasoline.unleaded.regular <- apiDF(json$Results$series[[9]]$data)
gasoline.unleaded.midgrade <- apiDF(json$Results$series[[10]]$data)
gasoline.unleaded.premium <- apiDF(json$Results$series[[11]]$data)


## Rename value prior to merging
names(electricity)[4]<- 'electricity'
names(utility.piped.gas.services)[4]<- 'utility.piped.gas.services'
names(new.used.motor.vehicles)[4]<- 'new.used.motor.vehicles'
names(new.vehicles)[4]<- 'new.vehicles'
names(used.cars.trucks)[4]<- 'used.cars.trucks'
names(motor.fuel)[4]<- 'motor.fuel'
names(gasoline.all.types)[4]<- 'gasoline.all.types'
names(motor.vehicle.insurance)[4]<- 'motor.vehicle.insurance'
names(gasoline.unleaded.regular)[4]<- 'gasoline.unleaded.regular'
names(gasoline.unleaded.midgrade)[4]<- 'gasoline.unleaded.midgrade'
names(gasoline.unleaded.premium)[4]<- 'gasoline.unleaded.premium'

## Merge data frames, convert cpi columns to numeric and create column for date

data_list <- list(
  electricity, utility.piped.gas.services, new.used.motor.vehicles, new.vehicles, used.cars.trucks,
  motor.fuel, gasoline.all.types, motor.vehicle.insurance, gasoline.unleaded.regular, gasoline.unleaded.midgrade,
  gasoline.unleaded.premium
)

cpi2 <- data_list %>% reduce(merge, all=T) %>% 
                      mutate_at(c(4:14), as.numeric) 


## Create new column for date
#cpi2 <- cpi2 %>% mutate(date = as.POSIXct(strptime(paste0('1',cpi2$periodName,cpi2$year),
#                                                   '%d%B%Y')))
#mutate(date = as.Date(with(cpi2, paste(year,periodName,'1',sep="-")),"%Y-%m-%d"))


## Clean environment
rm (
  electricity, utility.piped.gas.services, new.used.motor.vehicles, new.vehicles, used.cars.trucks,
  motor.fuel, gasoline.all.types, motor.vehicle.insurance, gasoline.unleaded.regular, gasoline.unleaded.midgrade,
  gasoline.unleaded.premium, data_list, json, payload, response, apiDF
)
```

```{r, include=FALSE, echo=FALSE}

## Merge the Two Data frames

cpi_use <- inner_join(cpi, cpi2)

rm(cpi, cpi2)
```


Below are the top and bottom five rows of the data.

### Top 10 rows
```{r}
cpi_use %>% head(10) %>% 
  kbl(align = "rrrrrrrrrrrrrrr") %>% 
  kable_material(c("striped", "hover"),
                 html_font = "calibri") %>% 
  scroll_box(width = "100%", height = "300px")
```


### Bottom 10 rows
```{r}
cpi_use %>% 
  tail(10) %>% 
  kbl(align = "rrrrrrrr") %>% 
  kable_material(c("striped", "hovered"),
                      html_font = "calibri") %>% 
  scroll_box(width = "100%", height = "300px")
```


